## Section 3.1

### Estimation of the Population Mean

To demonstrate properties of the sample mean as an estimator for the corresponding population mean consider the following code snippet.



<div class="unfolded">
```{r, echo = T, eval = T, message = F, warning = F}
pop <- rnorm(25000, 10, 1)
est <- replicate(expr = mean(sample(x = pop, size = 100)), n = 25000)
head(est)
```
</div>

We generate a population `x` which consists of 10000 observations drawn from the normal distribution with $\bar{x} = 10$ and $\hat{\sigma}^2 = 1$. Afterwards we use the function `replicate` to evaluate the code expression in `expr` $n = 25000$ times.
That is, we take 25000 samples of size 100 and compute the sample mean for each of them.

The plot below...

<div class="folded">
```{r, echo = T, eval = T, message = F, warning = F}
plot(density(est), col = 'steelblue', lwd = 2, bty = 'l')
abline(v = 10, lty = 2)
```
</div>









## Section 3.5


### Table 3.1: The Gender Gap of Earnings of College Graduates in the United States
In order to reproduce the data from table 3.1 you need to download the replication data hosted by Pearson (http://wps.aw.com/aw_stock_ie_3/178/45691/11696965.cw/index.html). Download the data for chapter three as an excel spreadsheet ("cps_ch3.xlsx") and load it in R with the following commands (make sure you use the correct path to the downloaded file). 

You need the packages *readxl* (`install.packages('readxl')`) to read the excel file and *dplyr* (`install.packages('dplyr')`) to make handling the data easier. 


```{r, echo=T, eval=T, message = F, warning = F}
library(readxl) # needed to read Excel files
library(dplyr, quietly = T)

cps <- read_excel(path = 'data/cps_ch3.xlsx')
# Get an overview of the data structure
head(cps)


# Take cps_dat %>% group data by gender and year %>% compute the mean, standard deviation
# and number of observations for each group
avgs <- cps %>% group_by(a_sex, year) %>% summarise(mean(ahe08), sd(ahe08), n())
print(avgs, digits = 2)
```

With the pipe operator `%>%` we simply chain different R commands. In the above code for example, we take the dataset `cps` and use it as an input for the function `group_by`. The output of that function will be used as an input for `summarise` and so on.

Now that we have computed the statistics of interest for both genders, we can investigate the gap between them.


```{r, echo=T}
male   <- avgs %>% filter(a_sex == 1) 
female <- avgs %>% filter(a_sex == 2)

# Rename columns
colnames(male)   <- c("Sex", "Year", "Y_bar_m", "s_m", "n_m")
colnames(female) <- c("Sex", "Year", "Y_bar_f", "s_f", "n_f")

# Gender gap, standard errors and confidence intervals
gap      <- male$Y_bar_m - female$Y_bar_f
gap_se   <- sqrt(male$s_m^2 / male$n_m + female$s_f^2 / female$n_f)
gap_ci_l <- gap - 1.96 * gap_se
gap_ci_u <- gap + 1.96 * gap_se
result   <- cbind(male[,-1], female[,-(1:2)], gap, gap_se, gap_ci_l, gap_ci_u)
print(result, digits = 2)
```
