Until now we assumed the regression function to be linear, i.e. the slope of the regression function to be constant. This implies that the effect on $Y$ of a unit change in $X$ does not depend on the value of $X$. If the effect of a change in $X$ on $Y$ depends on the value of $X$, we have to use a nonlinear regression function.

Let us have a look at an example where using a nonlinear regression function might be better suited to describe the relationship between $X$ and $Y$: the relation between `District Income` and `Test Scores` (Stock and Watson, 2012 - Figure 8.2).

```{r, results='hide', echo=TRUE, message=FALSE}
#Preparing the data
library(AER)                                                    # contains the dataset 
data(CASchools)
CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
```

We start our analysis by computing the correlation between the two variables. 

```{r}
cor(CASchools$income, CASchools$score)
```

The correlation coefficient is about `0.71. This means that income and test scores are positively correlated. In other words, children whose parents have an above average income tend to achieve above average test scores. Can we use the correlation coefficient to assess whether a linear regression model does fit the data adequately? To answer this question we visualize the data and add a linear regression line to the plot.

```{r, fig.align = 'center', fig.cap='Figure 8.1'}
#Fit linear model
linear_model<- lm(score ~ income, data = CASchools)

# Plot observations
plot(CASchools$income, CASchools$score,
     col = "steelblue",
     pch = 20,
     xlab="District Income (thousands of dollars)", 
     ylab="Test Score")

# Estimate a linear model and add the regression line to the plot
abline(linear_model, col="red", lwd=2)
```

As Stock and Watson point out, the linear regression line seems to overestimates the true relationship when income is very high or very low and underestimates it in the midrange.    

Fortunately, OLS is not restricted to linear functions. We can for example model `Test Scores` as a function of `income` and the square of `income`.

$$TestScore_i = \beta_0 + \beta_1 Income_i + \beta_2 Income_i^2 + u_i$$

This equation is called the *quadratic regression model*. Note, that $Income^2$ is treated as an additional explanatory variable. Hence, the quadratic model is a special case of a multivariate regression model. When fitting the model with `lm` we can use the `^` operator in addition with the function `I()` to add the quadratic term to our regression model. 

```{r, fig.align = 'center', fig.cap='Figure 8.2'}
#Fit Model
quadratic_model <- lm(CASchools$score ~ CASchools$income + I(CASchools$income^2))

#Model Output
summary(quadratic_model)
```

We might want to draw the same plot as for the linear model and add the regression line for the quadratic model. `abline()` can only draw straight lines and can not be used for this task. A function which can be used to draw lines but without being restricted to straight lines is
`lines(x_values, y_values)`. The x and y values serve as coordinates. The function draws points at the provided coordinate pairs and connects 
them sequentially. This makes it necessary to sort the coordinate pairs according to the X-values. Otherwise your will not get the desired result. 


```{r, fig.align="center"}
plot(CASchools$income, CASchools$score,
     col  = "steelblue",
     pch = 20,
     xlab = "District Income (thousands of dollars)",
     ylab = "Test Score")

#Add linear function to the plot
abline(linear_model , col="black", lwd=2)

#Add quatratic function to the plot
order_id <- order(CASchools$income)

lines(x = CASchools$income[order_id], 
      y = fitted(quadratic_model)[order_id],
      col="red", lwd=2) 

```


The approach we used to obtain a quadratic model can be generalized to polynomial models of arbitrary order. 
$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \ldots + \beta_r X_i^r + u_i$$

A cubic model for instance can be estimated in the same way as the quadratic model.

```{r cubic}
cubic_model <- lm(score ~ income + I(income^2) + I(income^3), data = CASchools)
```

In practice the question will arise which model should be chosen. For reasons already discussed in section ... 
the $R^2$ and the **SER** are not appropriate because adding regressors always improves those measures. However, we can
test whether the estimated coefficients are significantly different from zero.  

```{r}
library(lmtest)
coeftest(cubic_model, vcov = vcovHC(cubic_model, type = "HC1"))
```

The coefficients for `income^3` are significant at the $5\%$ level. This means we reject the hypothesis that the 
regression function is quadratic against the alternative that it is cubic. Furthermore, we could test if the coefficients 
for `income^2` and `income^3` are jointly significant. 


```{r f-test}
anova(linear_model, cubic_model, test = "F") ##Seite 307 F-statistic = 37.7?
```

With a $p$-value of `1.424e-10` i.e. much less than $0.05$, the null hypothesis that the regression function is linear is rejected inf favour of the alternative, that is a
quadratic or a cubic model. 

### Interpretation of coefficients in polynomial regression models 

The coefficients in polynomial regression do not have a simple interpretation. The best way is to calculate the estimated effect on $Y$ associated with a change in $X$ for one or more values of $X$. 

<div class = "keyconcept"> 
<h3 class = "right"> Key Concept 8.1 </h3>          
<h3 class = "left"> The Expected Effects on Y of a Change in $X_1$ in the Nonlinear Regression Model </h3>

<p> The expected change in $Y$, $\Delta Y$, associated with the change in $X_1$, $\Delta X_1$, holding $X_2, \cdots , X_k$ constant. That is, the expected change in $Y$ is the difference: 

$$\Delta Y = f(X_1 + \Delta X_1, X_2, \cdots, X_k) - f(X_1, X_2, \cdots, X_k).$$

The estimator of this unknown population difference is the difference between the predicted values for these two cases. Let $\hat{f}(X_1, X_2, \cdots, X_k)$ be the predicted value of of $Y$ based on the estimator $\hat{f}$ of the population regression function. Then the predicted change in $Y$ is

$$\Delta \hat{Y} = \hat{f}(X_1 + \Delta X_1, X_2, \cdots, X_k) - \hat{f}(X_1, X_2, \cdots, X_k).$$
</p>
</div>

### Application to test scores and income

What is the predicted change in test score associated with a one unit change ($\$1.000$), based on the estimated quadratic regression function

$$\widehat{TestScore} = 607.3 + 3.85 \times Income - 0.0423 \times Income^2 \text{ ?}$$

Because the regression function is quadratic, this effect depends on the initial district Income. We therefore consider two cases: an increase 
in district income form $10$ to $11$ (i.e. from $\$10,000$ per capita to $\$11,000$) and an increase in district income from $40$ to $41$. 

To compute compute $\Delta \hat{Y}$ associated with a change in income form $10$ to $11$, we use the following formula: 

$$\Delta \hat{Y} = \left(\hat{\beta_0 } + \hat{\beta_1} \times 11 + \hat{\beta_2} \times 11^2\right) - \left(\hat{\beta_0 } + \hat{\beta_1} \times 10 + \hat{\beta_2} \times 10^2\right) $$
To compute $\hat{Y}$ using R we apply `predict()`.

```{r quadratic}
quadriatic_model <- lm(score ~ income + I(income^2), data = CASchools)
new_data         <- data.frame(income = c(10,11,40,41))
Y_hat            <- predict(quadriatic_model,newdata = new_data)
changes          <- matrix(Y_hat, nrow = 2, byrow = TRUE) #Zu umständlich?
changes[ ,2] - changes[ ,1]
```

A change of $X$ from $10$ to $11$ increases the predicted score by $2.96$ whereas 
a change from $40$ to $41$ only increases the predicted score by $0.42$. Hence the slope of the estimated
quadratic regression function is steeper at low levels of income than at the higher levels. 



### Logarithms
Another way to specify a nonlinear regression function is to use the natural logarithm of $Y$ and/or $X$.
Logarithms convert changes in variables into percentage changes, and many relationships are naturally expressed in terms of percentages. 

There are three different cases in which logarithms might be used, when X is transformed by taking its logarithm but Y is not; when Y is transformed
to its logarithm but X is not; and when both Y and X are transformed to their logarithms. The interpretation of the regression coefficients is different in each case.  

#### Case I: x is in logarithm, Y is not.

The regression model is 

$$Y_i = \beta_0 + \beta_1 \times \ln(X_i) + u_i \text{, } i=1,...,n. $$
Similar as for polynomial regression we do not have to create a new variable $log(x)$. We can specify the formula argument 
of `lm()` to tell R that the log transformation of a variable should be used.    

```{r}
linearLog_model <- lm(score ~ log(income), data = CASchools)
coeftest(linearLog_model, vcov = vcovHC(linearLog_model, type = "HC1"))
```

According to the output the estimated regression function is:

$$\widehat{TestScore} = 557.8 + 36.42 \times \ln(Income).$$

```{r, fig.align="center"}
plot(score ~ income, col="steelblue", pch=20, data = CASchools)
order_id  <- order(CASchools$income)
lines(CASchools$income[order_id], fitted(linearLog_model)[order_id], col = "red", lwd = 2)
```


We can interpret $\beta_1$ as follows: a $1\%$ increase in income is associated with an increase in test scores of $0.01 \times 36.42 = 0.36$ points.
To estimate the effect on $Y$ in its original units of thousends of dollars (not in logarithms), the method from Key Concept 8.1 can be used.


```{r}
Y_hat <- predict(linearLog_model,newdata = new_data)
changes <- matrix(Y_hat, nrow = 2, byrow = TRUE)
changes[ ,2] - changes[ ,1]
```

#### Case II: Y is in logarithm, X is not

#### Case III: X and Y are in logarithm

```{r log log}
logLog_model <- lm(log(score) ~ log(income), data = CASchools)
coeftest(logLog_model, vcov = vcovHC(logLog_model, type = "HC1"))
```

```{r, fig.align="center"}
plot(log(score) ~ income, data = CASchools,
     col = "steelblue", pch=20)
lines(sort(CASchools$income), fitted(logLog_model)[order(CASchools$income)], col = "red", lwd = 2)
```

```{r}
Y_hat   <- predict(linearLog_model,newdata = new_data)
changes <- matrix(Y_hat, nrow = 2, byrow = TRUE)
changes[ ,2] - changes[ ,1]
```


<div class = "keyconcept"> 
<h3 class = "right"> Key Concept 8.2 </h3>          
<h3 class = "left"> Logarithms in Regression: Three Cases </h3>

<p> Logarithms can be used to transform the dependent variable $Y$, an independent variable $X$, or both 
(but the variable being transformed must be positive). The following table summarizes these three cases and the interpretation of the regression coefficient $\beta_1$. In each case, $\beta_1$, can be estimated applying OLS after taking the logarithm if the dependent and/or independent variable.

<table>
<thead>
<tr class="header">
<th align="left">Case</th>
<th align="left">Model Specification</th>
<th align="left">Interpretation of $\beta_1$</th>
</tr>
</thead>
<tbody>
<tr class="odd">
<td align="left">I</td>
<td align="left">$Y_i = \beta_0 + \beta_1 \ln{X_i} + u_i$</td>
<td align="left">A $1 \%$ change in $X$ is associated with a change in $Y$ of $0.01 \beta_1$.</td>
</tr>
<tr class="even">
<td align="left">II</td>
<td align="left">$\ln(Y_i) = \beta_0 + \beta_1 X_i + u_i$</td>
<td align="left">A change in $X$ by one unit ($\Delta X = 1$) is associated with a $100\beta_1 \%$ change in $Y$.</td>
</tr>
<tr class="odd">
<td align="left">III</td>
<td align="left">$\ln(Y_i) = \beta_0 + \beta_1 \ln(X_i) + u_i$</td>
<td align="left">A $1 \%$ change in $X$ is associated with a $100\beta_1 \%$ change in $Y$, so $\beta_1$ is the elasticity of $Y$ with respect to $X$.</td>
</tr>
</tbody>
</table>

</p>
</div>




```{r poly log}
polyLog_model <- lm(score ~ log(income) + I(log(income)^2) + I(log(income)^3), data = CASchools)
coeftest(polyLog_model, vcov = vcovHC(polyLog_model, type = "HC1"))

```
