---
title: "Introduction to Econometrics with R"
author: "M. Arnold, M. Schmelzer, A. Gerber"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<!--Include script for hiding output chunks-->
<script src="js/hideOutput.js"></script>


# Chapter 3
## Section 3.5

### Table 3.1 - Not completed (sig. test)
In order to reproduce the data from table 3.1 you need to download the data from the books source (LINK).
We simply read in the data and then make use of the powerful package dplyr to subset the data.

```{r, echo=T, eval=T}
library(dplyr, quietly=T)
# Read in CPS data (taken from http:// ....)
cps_dat <- read.csv("data/cps_ch3.csv", sep=";", dec=".")
# Get an overview of the data structure
head(cps_dat)
# Take cps_dat %>% group data by gender and year %>% compute the mean, standard deviation
# and number of observations for each group
avgs <- cps_dat %>% group_by(a_sex, year) %>% summarise(mean(ahe12), sd(ahe12), n())
print(avgs, digits=2)
```

Now that we have the numbers we can compute the difference between both genders.
```{r, echo=T}
male   <- avgs %>% filter(a_sex == 1) 
female <- avgs %>% filter(a_sex == 2)
colnames(male)   <- c("Sex", "Year", "Y_bar_m", "s_m", "n_m")
colnames(female) <- c("Sex", "Year", "Y_bar_f", "s_f", "n_f")

# Gender gap, standard errors and confidence intervals
gap      <- male$Y_bar_m - female$Y_bar_f
gap_se   <- sqrt(male$s_m^2 / male$n_m + female$s_f^2 / female$n_f)
gap_ci_l <- gap - 1.96 * gap_se
gap_ci_u <- gap + 1.96 * gap_se
result <- cbind(male[,-1], female[,-(1:2)], gap, gap_se, gap_ci_l, gap_ci_u)
print(result, digits = 2)
```


# Chapter 4
## Linear Regression with One Regressor 

This chapter introduces the linear regression model which  relates one variable, $X$, to another variable $Y$. If for example a school cuts the class sizes by hiring new teachers, would this have an impact on the students test scores? With linear regression we can not only examine whether the class size ($X$) does have an impact on the test results ($Y$). We can learn also something about the direction and the strength of the effect.      

```{r}
library(AER)                                                    # contains the dataset 
data(CASchools) 

CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
```


```{r Table 4.1, results='hold'}
mean(CASchools$tsratio) 
mean(CASchools$score)
sd(CASchools$tsratio) 
sd(CASchools$score)
quantiles          <- c(0.10, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9)
quantile(CASchools$tsratio, quantiles)
quantile(CASchools$score, quantiles)
```


```{r Figure 4.2}
plot(score ~ tsratio, 
     data = CASchools,
     main = "Scatterplot of Test Score vs. Student-Teacher Ratio", 
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     xlim = c(10,30),
     ylim = c(600, 720))
```

## The Ordinary Least Squares Estimator (OLS)

\begin{align}
\hat{\beta_1} & =  \frac{\sum_{i = 1}^n (X_i - \bar{X})(Y_i - \bar{Y})} { \sum_{i=1}^n (X_i - \bar{X})^2} \\

\hat{\beta_0} & =  \bar{Y} - \hat{\beta_1} \bar{X} 
\end{align}




```{r}
attach(CASchools)
beta_1 <- sum((tsratio - mean(tsratio))*(score - mean(score))) / sum((tsratio - mean(tsratio))^2)
beta_0 <- mean(score) - beta_1 * mean(tsratio)
```

```{r}
plot(score ~ tsratio, 
     data = CASchools,
     main = "Scatterplot of Test Score vs. Student-Teacher Ratio", 
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     xlim = c(10,30),
     ylim = c(600, 720))

linear_model <- lm(score ~ tsratio, data = CASchools)
linear_model

abline(a = beta_0, b = beta_1)
```


```{r}
linear_model <- lm(score ~ tsratio, data = CASchools)
linear_model
```

```{r Figure 4.3}
plot(score ~ tsratio, 
     data = CASchools,
     main = "Scatterplot of Test Score vs. Student-Teacher Ratio", 
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     xlim = c(10,30),
     ylim = c(600, 720))

abline(linear_model)
```

# Chapter 5

## Section 5.1

Consider again the previous OLS regression that gave us the regression line
  \[ \widehat{score} = 698.9 - 2.28 \times tsratio.  \]
For testing a hypothesis regarding the slope parameter, we need $SE(\hat{\beta}_1)$, the standard error of the respective point estimator. By looking at the second column of the coefficient's summary, we find this value to be approximately $0.48$.

```{r}
summary(linear_model)$coefficients
```

How is $SE(\hat{\beta}_1)$ computed? The `summary` function uses the homoskedasticity-only formula 
  \[ SE(\hat{\beta}_1) = \sqrt{ \overset{\sim}{\sigma}^2_{beta_1} } = \sqrt{ \frac{SER^2}{\sum(X_i - \overline{X})^2} }. \]


## Section 5.4

```{r Figure 5.2}
library(scales)

# Genrate some heteroscedastic data

set.seed(123) 
x <- rep(c(10,15,20,25),each=25)
e <- rnorm(100, sd=12)                
i <- order(runif(100, max=dnorm(e, sd=12))) 
y <- 720 - 3.3 * x + e[rev(i)]

# Plot data and inspect conditional distributions

mod <- lm(y ~ x)
plot(x, y, main="An Example of Heteroscedasticity",
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     cex=0.5, pch=19, xlim=c(8,27), ylim=c(600,710))
abline(mod, col="red")
boxplot(y~x, add=TRUE, at=c(10,15,20,25), col=alpha("gray", 0.4), border="black")
```

```{r}
#if(!require(devtools))
#  install.packages("devtools")
#library(devtools)
#install_github("datacamp/tutorial")
#library(tutorial)
```

We already have assigned the model's summary to `mod`. Now, use your R knowledge to display coefficients and standard errors!

```{r, ex="create_a", type="pre-exercise-code", include=FALSE}
# Available in the workspace when the session initializes
set.seed(123) 
x <- rep(c(10,15,20,25),each=25)
e <- rnorm(100, sd=12)                
i <- order(runif(100, max=dnorm(e, sd=12))) 
y <- 720 - 3.3 * x + e[rev(i)]
mod <- summary(lm(y ~ x))
```

```{r, ex="create_a", type="sample-code", include=FALSE}
# Access coefficients and standard errors

```

```{r, ex="create_a", type="solution", include=FALSE}
# Access coefficients and standard errors
mod$coefficients
```

```{r, ex="create_a", type="sct", eval=FALSE, include=FALSE}
test_output_contains("mod$coefficients", incorrect_msg = "Make sure to use `mod$` !")
success_msg("Econometrics ROXXXXXX111!!!")
```

```{r, ex="create_a", type="hint", eval=FALSE, include=FALSE}
Use the dollar operator.
```

# Chapter 6
## Section 6.3
```{r}
# Multiple Regression - Application to Test Scores and the Student-Teacher Ratio
library(AER)                                                    # contains the dataset 
data(CASchools) 
CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
model <- lm(score ~ tsratio + english, data = CASchools)
```

## Section 6.4
Taking the code from Section 6.3 you could simply use `summary(model)` to print the $SER$, $R^2$ and adjusted-$R^2$. You find them at the bottom of the output.
```{r}
summary(model)
```
You can also compute the parameters by hand using the formulas on p. 242.
```{r}
n <- nrow(CASchools)    # number of observations (rows)
k <- 2                  # 2 regressors
y_mean <- mean(CASchools$score)

SSR <- sum( residuals(model)^2 )
TSS <- sum( ( CASchools$score - y_mean )^2 )
ESS <- sum((fitted(model) - y_mean)^2)

SER <- sqrt(1/(n-k-1) * SSR)
Rsq <- 1 - (SSR / TSS)
adj_Rsq <- 1 - (n-1)/(n-k-1) * SSR/TSS
cat(SER, Rsq, adj_Rsq, sep = "    ")
```

## Section 6.7
### Examples of Perfect Multicollinearity

If you use `lm` to estimate a model with a set of regressors that suffer from perfect multicollinearity the system will warn you (_1 not defined because of singularities_) and ignore the regressor(s) which is(are) assumed to be a linear combination of the others.
```{r}
# Example 1
library(AER)                                                    # contains the dataset 
data(CASchools) 
CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
CASchools$FracEL    <- CASchools$english/100
model <- lm(score ~ tsratio + english + FracEL, data = CASchools)
summary(model)
```


If you compute OLS by hand you will run into the problem as well but noone is helping you out. The computation simply fails. Why is this the case? Take the following example:

Assume you want to estimate a simple linear regression model with an intercept and one regressor:
    \[ y_i = \beta_0 + \beta_1 x_i + u_i\]
When applying perfect multicollinearity $x$ has to be a linear combination of the other regressors. Since the only other regressor is a constant, $x$ has to be constant as well. When you recap the formula for $\beta_1$, namely
    \[ \hat{\beta_1} =  \frac{\sum_{i = 1}^n (X_i - \bar{X})(Y_i - \bar{Y})} { \sum_{i=1}^n (X_i - \bar{X})^2}\]
you find the variance of the regressor $x$ in the nominator. Since the variance of a constant is zero, you are not able to compute this fraction. $\hat{\beta}_1$ remains undefined.

<font style="color:#004c93; font-weight:bold;">Note:</font> In this special case the nominator equals zero as well. Can you show that?


```{r}
# Example 2
library(AER)                                                    # contains the dataset 
data(CASchools) 
CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
CASchools$NVS      <- ifelse(CASchools$tsratio < 12, 0, 1)      # if tsratio smaller 12, NVS = 0, else NVS = 1
model <- lm(score ~ tsratio + english + NVS, data = CASchools)
summary(model)
```


```{r}
# Example 3
library(AER)                                                    # contains the dataset 
data(CASchools) 
CASchools$tsratio  <- CASchools$students/CASchools$teachers  # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2  # average test-score
CASchools$PctES    <- 100 - CASchools$english  # Percentage of english speakers 
model <- lm(score ~ tsratio + english + PctES, data = CASchools)
summary(model)
```



