---
title: "Introduction to Econometrics with R"
author: "M. Arnold, M. Schmelzer, A. Gerber"
date: "20 Mai 2016"
output:
  html_document:
    css: styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Chapter 3
## Section 3.5

### Table 3.1 - Not completed (sig. test)
In order to reproduce the data from table 3.1 you need to download the data from the books source (LINK).
We simply read in the data and then make use of the powerful package dplyr to subset the data.

```{r, echo=T, eval=T}
library(dplyr, quietly=T)
# Read in CPS data (taken from http:// ....)
cps_dat <- read.csv("data/cps_ch3.csv", sep=";", dec=".")
# Get an overview of the data structure
head(cps_dat)
# Take cps_dat %>% group data by gender and year %>% compute the mean, standard deviation
# and number of observations for each group
avgs <- cps_dat %>% group_by(a_sex, year) %>% summarise(mean(ahe12), sd(ahe12), n())
print(avgs, digits=2)
```

Now that we have the numbers we can compute the difference between both genders.
```{r, echo=T}
male   <- avgs %>% filter(a_sex == 1) 
female <- avgs %>% filter(a_sex == 2)
colnames(male)   <- c("Sex", "Year", "Y_bar_m", "s_m", "n_m")
colnames(female) <- c("Sex", "Year", "Y_bar_f", "s_f", "n_f")

# Gender gap, standard errors and confidence intervals
gap      <- male$Y_bar_m - female$Y_bar_f
gap_se   <- sqrt(male$s_m^2 / male$n_m + female$s_f^2 / female$n_f)
gap_ci_l <- gap - 1.96 * gap_se
gap_ci_u <- gap + 1.96 * gap_se
result <- cbind(male[,-1], female[,-(1:2)], gap, gap_se, gap_ci_l, gap_ci_u)
print(result, digits = 2)
```


# Chapter 4

```{r}
library(AER)                                                    # contains the dataset 
data(CASchools) 

CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
```


```{r Table 4.1}
mean(CASchools$tsratio) 
mean(CASchools$score)
sd(CASchools$tsratio) 
sd(CASchools$score)
quantiles          <- c(0.10, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9)
quantile(CASchools$tsratio, quantiles)
quantile(CASchools$score, quantiles)
```


```{r Figure 4.2}
plot(score ~ tsratio, 
     data = CASchools,
     main = "Scatterplot of Test Score vs. Student-Teacher Ratio", 
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     xlim = c(10,30),
     ylim = c(600, 720))
```

```{r}
attach(CASchools)
beta_1 <- sum((tsratio - mean(tsratio))*(score - mean(score))) / sum((tsratio - mean(tsratio))^2)
beta_0 <- mean(score) - beta_1 * mean(tsratio)
```

```{r}
plot(score ~ tsratio, 
     data = CASchools,
     main = "Scatterplot of Test Score vs. Student-Teacher Ratio", 
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     xlim = c(10,30),
     ylim = c(600, 720))

linear_model <- lm(score ~ tsratio, data = CASchools)
linear_model

abline(a = beta_0, b = beta_1)
```


```{r}
linear_model <- lm(score ~ tsratio, data = CASchools)
linear_model
```

```{r Figure 4.3}
plot(score ~ tsratio, 
     data = CASchools,
     main = "Scatterplot of Test Score vs. Student-Teacher Ratio", 
     xlab = "Student teacher-ratio (X)",
     ylab = "Test Score (Y)",
     xlim = c(10,30),
     ylim = c(600, 720))

abline(linear_model)
```


Änderung 

# Chapter 5

# Chapter 6
## Section 6.3
```{r}
# Multiple Regression - Application to Test Scores and the Student-Teacher Ratio
library(AER)                                                    # contains the dataset 
data(CASchools) 
CASchools$tsratio  <- CASchools$students/CASchools$teachers     # teacher-student-ratio
CASchools$score    <- (CASchools$read + CASchools$math)/2       # average test-score
model <- lm(score ~ tsratio + english, data = CASchools)
```

## Section 6.4
Taking the code from Section 6.3 you could simply use `summary(model)` to print the $SER$, $R^2$ and adjusted-$R^2$. You find them at the bottom of the output.
```{r}
summary(model)
```
You can also compute the parameters by hand using the formulas on p. 242.
```{r}
n <- nrow(CASchools)    # number of observations (rows)
k <- 2                  # 2 regressors
y_mean <- mean(CASchools$score)

SSR <- sum( residuals(model)^2 )
TSS <- sum( ( CASchools$score - y_mean )^2 )
ESS <- sum((fitted(model) - y_mean)^2)

SER <- sqrt(1/(n-k-1) * SSR)
Rsq <- 1 - (SSR / TSS)
adj_Rsq <- 1 - (n-1)/(n-k-1) * SSR/TSS
cat(SER, Rsq, adj_Rsq, sep = "    ")
```




